---
title: Inspecting Konkor 1397 (2018) Data
author: Masood Sadat
date: '2018-08-24'
slug: inspecting-konkor-1397-data
categories:
  - afghanistan
  - custom Analysis
tags:
  - visualization
draft: true
---

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
# library(highcharter)
# library(tidyverse)
# library(plotly)
```


```{r}
library(tidyverse)
library(treemap)
library(data.table)
library(knitr)

konkor <- readRDS("~/Documents/PRSO/Programs/Data Analysis with R/week 4/konkor/konkor.RDS")

data <- setDT(konkor)[, list(n = .N, score = mean(score, na.rm = T)), by = province]

treemap(data,
        index = "province",
        vSize = "n",
        vColor = "score",
        type="manual",
        palette=c("#FFFFFF", "#6C512E"),
        title = "")
```

> Read about the **Konkor** exam in Afghanistan in the footnote.^[The nationwide university entrance exam for public universities is called Konkor exam. Every year, tens of thousands of students from all over the country participate in this exam. Top students usually spend years in preparation for this exam. Parents put a lot expectations on their children to perform well and get admission to top faculties and universities. To some extent, success becomes a family pride and failure becomes shame. Whether students prepare a lot or none, taking Konkor exam is likely the most stressful experience of their lives up to that point (around 18 years old). I have taken this exam in 2006 with relatively good score that admitted me to the Engineering faculty of Balkh University. However, I studied only one semester there before moving to India to study my undergraduate education.]

**Disclaimer:** Konkor data is not public, but Konkor results dataset was publicly shared by various individuals on social media on August 1, 2018. As the dataset include personal information, I will aggregate the results at provincial level without using personal information such as names.

The data has the following variables or indicators:

```{r}
data.table(names(konkor), 
           c("Name of test taker", "Gender of test taker",
             "Year graduated from high school", "Konkor test score",
             "Name of Province", "Whether test taker was caught cheating",
             "Whether test taker failed or admitted to university",
             "Whether test taker graduated from a private university",
             "Not admitted but eligible for Private Universities")) %>% 
  kable(col.names = c("Variables", "Description"))
```

## Participation

Overall, approximately `r round(nrow(konkor)/1000)` thousand high school graduates took part in the konkor exam of 1397, from all provinces. Larger and more populated provinces had also larger number of test takers, with the exception of Kandahar province where only `r konkor[province=="Kandahar", .N]` students took part--ranking 21 in terms of number of test takers. Kabul on the other hand had the largest number of test takers by far (`r konkor[province=="Kabul", .N]` test takers). Graph below shows the top 10 provinces with the highest number of test takers.

```{r}
test.takers <- konkor[, .N, by = province][order(-N)]
ggplot(test.takers[1:10], aes(reorder(province, -N), N, fill = ordered(rev(N)))) +
  geom_col() +
  geom_text(aes(label = prettyNum(N, ",")), size = 4, vjust = -.5) + 
  ylim(0, max(test.takers$N)+.08*max(test.takers$N)) +
  theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = .5)) +
  labs(x = "", y = "", title = "Number of Test Takers by Province")
```


I am going to compare the number of test takers by province to the population estimates of those provinces. This way, I want to make a simplified conclusion where education faces more challenge than others. The population estimates come from the Central Statistics Organization of Afghanistan.^[Link to download population estimates: ]

```{r}
pop <- data.table(province = c("Kabul", "Kapisa", "Parwan", "Wardak", "Logar",
                               "Nangarhar", "Laghman", "Panjshir", "Baghlan",
                               "Bamyan", "Ghazni", "Paktika", "Paktia", "Khost",
                               "Kunar", "Nooristan", "Badakhshan", "Takhar",
                               "Kunduz", "Samangan", "Balkh", "Sar-e-Pul", "Ghor",
                               "Daikundi", "Uruzgan", "Zabul", "Kandahar",
                               "Jawzjan", "Faryab", "Helmand", "Badghis", "Herat",
                               "Farah", "Nimroz"),
                  pop = c(4860880, 471574, 711621, 637634, 419377, 1635872, 476537,
                          164115, 977297, 478424, 1315041, 748910, 590668, 614584,
                          482115, 158211, 1017499, 1053852, 1091116, 415343,
                          1442847, 599137, 738224, 498840, 420964, 371043, 1337183,
                          579833, 1069540, 1395514, 530574, 2050514, 543237,
                          176898))
setorder(pop, -pop)
pop <- mutate(pop, order = 1:34) %>% 
  merge(test.takers) %>% arrange(order) %>% select(3, 1, 2, 4)
kable(mutate(pop, pop = prettyNum(pop, ",")), 
      col.names = c("Order", "Province", "Population Estimate", "Test Takers"))
```


```{r}
pop <- as.data.table(pop)
pop[, ratio := (N/pop)*10000]
setorder(pop, -ratio)
pop
ggplot(pop, aes(reorder(province, ratio), as.integer(ratio))) + 
  geom_col() +
  coord_flip()
```



## Performance




