---
title: Essentials of data.table
author: Masood Sadat
date: '2019-03-24'
slug: essentials-of-data-table
categories:
  - r tips
tags:
  - r tips
draft: true
---

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

## Motivation

**data.table** is a R package that offers an enhanced version of data.frame and allows very fast data manupulation. As data.table is data.frame too, all functions and operations that apply to data.frame also applies to data.table. However, there are many enhancements that are only available in data.table, and unavailable in data.frame. I used to be an avid user of [tidyverse](https://www.tidyverse.org) since I had started working with R. But recently, I started working with very large dataframes (few million obs. at the least) and there I had to change to **data.table** because of speed. After working for a few weeks, I started appreciating its concise syntax as well. Therefore, I decided to write this blog post to share my knowledge with others and for my future use.


## Subsetting

Using hard bracket notation **[** and **]**, I can subset a data.table object's rows or columns. Let's get started:

```{r}
# load data.table package
library(data.table)

# check class of df
df <- mtcars
class(df)

# turn df into data.table
setDT(df)
class(df)

# subset first five rows
df[1:5]

# subset columns
df[, 1:5]

# subset rows and columns
df[1:5, 1:5]

# subset by condition
df[mpg > 30]
df[mpg > 30 & wt > 2]
df[cyl %in% c(4, 8) & hp > 100]
```














